---
output:
  md_document:
    variant: markdown_github
html_preview: false
---

#  ggtree: an R package for visualization and annotation of phylogenetic trees with their covariates and other associated data

```{r echo=FALSE, results="hide", message=FALSE}
library("txtplot")
library("ypages")
```

`r badge_release("ggtree", "green")`
`r badge_devel("ggtree", "green")`
[![Bioc](http://www.bioconductor.org/shields/years-in-bioc/ggtree.svg)](https://www.bioconductor.org/packages/devel/bioc/html/ggtree.html#since)
`r badge_download("ggtree", "total", "blue")`
`r badge_download("ggtree", "month", "blue")`


[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
[![codecov](https://codecov.io/gh/GuangchuangYu/ggtree/branch/master/graph/badge.svg)](https://codecov.io/gh/GuangchuangYu/ggtree)
[![Last-changedate](https://img.shields.io/badge/last%20change-`r gsub('-', '--', Sys.Date())`-green.svg)](https://github.com/GuangchuangYu/ggtree/commits/master)
[![GitHub forks](https://img.shields.io/github/forks/GuangchuangYu/ggtree.svg)](https://github.com/GuangchuangYu/ggtree/network)
[![GitHub stars](https://img.shields.io/github/stars/GuangchuangYu/ggtree.svg)](https://github.com/GuangchuangYu/ggtree/stargazers)
[![Awesome](https://cdn.rawgit.com/sindresorhus/awesome/d7305f38d29fed78fa85652e3a63e154dd8e8829/media/badge.svg)](https://awesome-r.com/#awesome-r-graphic-displays)

[![platform](http://www.bioconductor.org/shields/availability/devel/ggtree.svg)](https://www.bioconductor.org/packages/devel/bioc/html/ggtree.html#archives)
[![Build Status](http://www.bioconductor.org/shields/build/devel/bioc/ggtree.svg)](https://bioconductor.org/checkResults/devel/bioc-LATEST/ggtree/)
[![Linux/Mac Travis Build Status](https://img.shields.io/travis/GuangchuangYu/ggtree/master.svg?label=Mac%20OSX%20%26%20Linux)](https://travis-ci.org/GuangchuangYu/ggtree)
[![AppVeyor Build Status](https://img.shields.io/appveyor/ci/Guangchuangyu/ggtree/master.svg?label=Windows)](https://ci.appveyor.com/project/GuangchuangYu/ggtree)
[![install with bioconda](https://img.shields.io/badge/install%20with-bioconda-green.svg?style=flat)](http://bioconda.github.io/recipes/bioconductor-ggtree/README.html)



The `ggtree` package extending the `ggplot2` package. It based on grammar of graphics and takes all the good parts of `ggplot2`. `ggtree` is designed for not only viewing phylogenetic tree but also displaying annotation data on the tree.

[![Twitter](https://img.shields.io/twitter/url/https/github.com/GuangchuangYu/ggtree.svg?style=social)](https://twitter.com/intent/tweet?hashtags=ggtree&url=http://onlinelibrary.wiley.com/doi/10.1111/2041-210X.12628/abstract&screen_name=guangchuangyu)

-----------------------------------------------------

Please cite the following article when using `ggtree`:

__G Yu__, DK Smith, H Zhu, Y Guan, TTY Lam^\*^. ggtree: an R package for visualization and annotation of phylogenetic trees with their covariates and other associated data. __*Methods in Ecology and Evolution*__. _accepted_

`r badge_doi("10.1111/2041-210X.12628", "green")`
`r badge_citation("HtEfBTGE9r8C", "7268358477862164627", "green")`
`r badge_altmetric("10533079", "green")`

----------------------------------------------------------------------------------------

For details, please visit our project website, <https://guangchuangyu.github.io/ggtree>.

+ [Documentation](https://guangchuangyu.github.io/ggtree/documentation/)
+ [FAQ](https://guangchuangyu.github.io/ggtree/faq/)
+ [Featured Articles](https://guangchuangyu.github.io/ggtree/featuredArticles/)
+ [Feedback](https://guangchuangyu.github.io/ggtree/#feedback)

### Citation

`r badge_citation("HtEfBTGE9r8C", "7268358477862164627", "green")`


```{r echo=F, comment=NA}
d <- get_article_cite_history('DO5oG40AAAAJ', 'HtEfBTGE9r8C')
txtplot(d$year, d$cites, xlim=c(2015, 2017))
```

### Download stats

[![download](http://www.bioconductor.org/shields/downloads/ggtree.svg)](https://bioconductor.org/packages/stats/bioc/ggtree/)
`r badge_download("ggtree", "total", "blue")`
`r badge_download("ggtree", "month", "blue")`


```{r echo=F, comment=NA}
x <- dlstats::bioc_stats("ggtree")
x <- x[-nrow(x), ] # last month data is not complete
txtplot(ggtree::Date2decimal(x$end), x$Nb_of_downloads, width=100, height=25)
```





