---
title: "ggtree: a phylogenetic tree viewer for different types of tree annotations"
author: "Guangchuang Yu and Tommy Tsan-Yuk Lam\\

        School of Public Health, The University of Hong Kong"
date: "`r Sys.Date()`"
bibliography: ggtree.bib
biblio-style: apalike
output:
  prettydoc::html_pretty:
    toc: true
    theme: cayman
    highlight: github
  pdf_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{00 ggtree introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r style, echo=FALSE, results="asis", message=FALSE}
knitr::opts_chunk$set(tidy = FALSE,
		   message = FALSE)
```


```{r echo=FALSE}
CRANpkg <- function (pkg) {
    cran <- "https://CRAN.R-project.org/package"
    fmt <- "[%s](%s=%s)"
    sprintf(fmt, pkg, cran, pkg)
}

Biocpkg <- function (pkg) {
    sprintf("[%s](http://bioconductor.org/packages/%s)", pkg, pkg)
}

```

> You can't even begin to understand biology, you can't understand life, unless you understand what it's all there for, how it arose - and that means evolution.
> --- Richard Dawkins


# Citation

If you use `r Biocpkg('ggtree')` in published research, please cite:

__G Yu__, DK Smith, H Zhu, Y Guan, TTY Lam^\*^. ggtree: an R package for
visualization and annotation of phylogenetic trees with their covariates and
other associated data. __*Methods in Ecology and Evolution*__. 2017, 8(1):28-36.
doi: [10.1111/2041-210X.12628](https://doi.org/10.1111/2041-210X.12628).


# Introduction

This project arose from our needs to annotate nucleotide substitutions in the
phylogenetic tree, and we found that there is no tree visualization software can
do this easily. Existing tree viewers are designed for displaying phylogenetic
tree, but not annotating it. Although some tree viewers can displaying bootstrap
values in the tree, it is hard/impossible to display other information in the
tree. Our first solution for displaying nucleotide substituitions in the tree is
to add this information in the node/tip names and use traditional tree viewer to
show it. We displayed the information in the tree successfully, but we believe
this indirect approach is inefficient.

Previously, phylogenetic trees were much smaller. Annotation of phylogenetic
trees was not as necessary as nowadays much more data is becomming available. We
want to associate our experimental data, for instance antigenic change, with the
evolution relationship. Visualizing these associations in a phylogenetic tree
can help us to identify evolution patterns. We believe we need a next generation
tree viewer that should be programmable and extensible. It can view a
phylogenetic tree easily as we did with classical software and support adding
annotation data in a layer above the tree. This is the objective of developing
the `r Biocpkg('ggtree')` [@yu_ggtree:_2017]. Common tasks of annotating a phylogenetic tree should
be easy and complicated tasks can be possible to achieve by adding multiple
layers of annotation.

The `r Biocpkg('ggtree')` is designed by extending the `r CRANpkg('ggplot2')`
[@wickham_ggplot2_2009] package. It is based on the grammar of graphics and
takes all the good parts of `r CRANpkg('ggplot2')`. There are other R packages that implement
tree viewer using `r CRANpkg('ggplot2')`, including `r CRANpkg('OutbreakTools')`,
`r Biocpkg('phyloseq')` [@mcmurdie_phyloseq_2013]
and [ggphylo](https://github.com/gjuggler/ggphylo); they mostly create complex
tree view functions for their specific needs. Internally, these packages
interpret a phylogenetic as a collection of lines, which makes it hard to
annotate diverse user input that are related to node (taxa). The `r Biocpkg('ggtree')` is
different to them by interpreting a tree as a collection of taxa and allowing
general flexibilities of annotating phylogenetic tree with diverse types of user
inputs.


# Getting data into *R*

Most of the tree viewer software (including *R* packages) focus on *Newick* and
*Nexus* file format, while there are file formats from different evolution
analysis software that contain supporting evidences within the file that are
ready for annotating a phylogenetic tree. The `r Biocpkg('treeio')` package
supports several file formats and software outputs. It brings analysis findings
to *R* users for further analysis (*e.g.* summarization, visualization,
comparison and test, *etc.*). It also allows external data to be mapped on the
phylogeny. Please refer to the `r Biocpkg('treeio')` vignette
for more details.

Users can use the following command to open the vignette:

```r
vignette("Importer", package="treeio")
```

All the data parsed/integrated by `r Biocpkg('treeio')` package can be used to
visualize or annotate phylogenetic tree in `r Biocpkg('ggtree')` [@yu_ggtree:_2017].


# Tree Visualization and Annotation

Tree Visualization in `ggtree` is easy, with one line of command `ggtree(tree_object)`. It supports several layouts, including `rectangular`, `slanted` and `circular` for `Phylogram` and `Cladogram`, `unrooted` layout, time-scaled and two dimentional phylogenies. [Tree Visualization](treeVisualization.html) vignette describes these feature in details.

We implement several functions to manipulate a phylogenetic tree.

+ taxa can be clustered together using `groupClade` or `groupOTU` functions
+ clades can be collapsed via `collapse` function
+ collapsed clade can be expanded by using `expand` function
+ clade can be re-scale to zoom in or zoom out by `scaleClade` function
+ selected clade can be rotated by 180 degree using `rotate` function
+ position of two selected clades (should share a same parent) can be exchanged by `flip` function

Details and examples can be found in [Tree Manipulation](treeManipulation.html) vignette.


Most of the phylogenetic trees are scaled by evolutionary distance (substitution/site), in `ggtree` a phylogenetic tree can be re-scaled by any numerical variable inferred by evolutionary analysis (e.g. species divergence time, *dN/dS*, _etc_). Numerical and category variable can be used to color a phylogenetic tree.

The `ggtree` package provides several layers to annotate a phylogenetic tree, including:

+ `geom_cladelabel` for labelling selected clades
+ `geom_hilight` for highlighting selected clades
+ `geom_range` to indicate uncertainty of branch lengths
+ `geom_strip` for adding strip/bar to label associated taxa (with optional label)
+ `geom_taxalink` for connecting related taxa
+ `geom_tiplab` for adding tip labels
+ `geom_treescale` for adding a legend of tree scale


It supports annotating phylogenetic trees with analyses obtained from R packages and other commonly used evolutionary software. User's specific annotation (e.g. experimental data) can be integrated to annotate phylogenetic trees. `ggtree` provides `write.jplace` function to combine Newick tree file and user's own data to a single `jplace` file that can be parsed and the data can be used to annotate the tree directly in `ggtree`.

`ggtree` integrates `phylopic` database and silhouette images of organisms can be downloaded and used to annotate phylogenetic directly. `ggtree` also supports using local images to annotate a phylogenetic tree.

Visualizing an annotated phylogenetic tree with numerical matrix (e.g. genotype table), multiple sequence alignment and subplots are also supported in `ggtree`. Examples of annotating phylogenetic trees can be found in the [Tree Annotation](treeAnnotation.html) and [Advance Tree Annotation](advanceTreeAnnotation.html) vignettes.


# Vignette Entry

+ [Tree Data Import](https://bioconductor.org/packages/devel/bioc/vignettes/treeio/inst/doc/Importer.html)
+ [Tree Visualization](treeVisualization.html)
+ [Tree Manipulation](treeManipulation.html)
+ [Tree Annotation](treeAnnotation.html)
+ [Advance Tree Annotation](advanceTreeAnnotation.html)
+ [Phylomoji](https://cran.r-project.org/web/packages/emojifont/vignettes/phylomoji.html)
+ [Annotating phylogenetic tree with images using ggtree and ggimage](https://cran.r-project.org/web/packages/ggimage/vignettes/ggtree.html)


More documents can be found in <https://guangchuangyu.github.io/ggtree>.

## Feedback ##

 - For bugs or feature request, please post to [github issue](https://github.com/GuangchuangYu/ggtree/issues).
 - For user questions, please post to [google group](https://groups.google.com/forum/#!forum/bioc-ggtree).


# Session info

Here is the output of `sessionInfo()` on the system on which this document was compiled:

```{r echo=FALSE}
sessionInfo()
```


# References
