---
title: "__*ggtree*__<br>phylogenetic tree viewer and annotator"
author: "Guangchuang Yu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggtree: phylogenetic tree viewer and annotator}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r echo=F}
library(ggplot2)
```

`ggtree` extending the `ggplot2` package to support phylogenetic tree object including phylo, ...
```{r fig.width=6, tidy=TRUE, width=60}
nwk <- "(((((((A:4,B:4):6,C:5):8,D:6):3,E:21):10,((F:4,G:12):14,H:8):13):13,((I:5,J:2):30,(K:11,L:11):2):17):4,M:56);"
library(ape)
tree <- read.tree(text=nwk)
library(ggtree)
ggplot(tree, aes(x, y)) + geom_tree() + theme_tree() + xlab("") + ylab("")
```

It's very easy to view a phylogenetic tree using `ggtree`, I have create a short cut, `ggtree` function, that works exactly the same as shown above. 
```{r eval=F}
ggtree(tree)
```

It works in the way you used `ggplot2`, for example, we can change the color, size and type of the lines.
```{r fig.width=6}
ggtree(tree, color="steelblue", size=1, linetype="dotted")
```

To show evolution distance, we can use `theme_tree2()` or `ggtree(showDistance=TRUE)`

```{r fig.width=6}
ggtree(tree) + theme_tree2()
```


Show all the internal nodes and tips in the tree can be done by adding a layer of points using `geom_point`.

```{r fig.width=6}
ggtree(tree) + geom_point()
ggtree(tree)+geom_point(aes(shape=isTip, color=isTip), size=3)
```
 
```{r fig.width=6}
ggtree(tree) + geom_tiplab()
ggtree(tree) + geom_tiplab(size=6, color="purple")
ggtree(tree) + geom_tiplab(align=TRUE, color="darkgreen") + geom_aline(color="steelblue", size=0.3)
```

```{r fig.width=6, fig.height=5}
ggtree(tree) + scale_x_continuous(trans="reverse")
ggtree(tree) + coord_polar(theta = "y")
```

## Tree annotation
Suppose we have the following data for annotation.
```{r}
dd <- data.frame(taxa=sample(LETTERS[1:13], 6), value=round(abs(rnorm(6))*100))
```
```{r echo=FALSE, results='asis'}
knitr::kable(dd)
```

```{r fig.width=6, fig.height=5}
ggtree(tree) + geom_tiplab() + geom_place(data=dd, map="taxa", place="value")
```

As the annotation overlap with tip labels, we can adjust the annotation position.

```{r fig.width=6, fig.height=5}
ggtree(tree) + geom_tiplab() + 
	     geom_place(data=dd, map="taxa", place="value", 
	     	        hjust=.8, vjust=-.4, size=4, color="red")
```





